Variability in upwelling patterns across the Benguela

### 2019-10-12 Write_up/Paper1_writeup.Rmd
fixing
updated_fixed_refs

Upwelling is a physical oceanographic process in which cold, nutrient rich waters are pushed to the surface as a result of the motion caused by alongshore equatorial winds. These upwelling events are known to have a large influence on the abundance, diversity, and distribution of marine organisms, and moreover, critically affect marine ecosystem processes (Add in references). Occurring globally throughout the world's oceans, four major eastern boundary upwelling systems (EBUS) can be recognised; each significantly impacting associated coastal systems. These respective systems are characterized as vast regions of coastal ocean occurring along the western shores of continents bordering the Pacific and Atlantic Oceans [@bakun; @pauly]. There has been significant interest in major upwelling systems since the later 1960s, with major interest occuring off Peru [@walsh] and North-West Africa [@barber]. Research done by Copenhagen 1953, reported a number of sites along the west coast of southern Africa where upwelling was particularly intense. The two sites of intense upwelling activity were found off the Cape Peninsula and off Cape columbine [@copenhagen; @andrew]. Whilst upwelling plays a significant role in ecological processes, climate change largely influences this phenomenon.

@@

Several studies have provided significant evidence suggesting that there has been an increase in ocean surface temperatures over the last few decades (Add references; RWS: You can cite the IPCC for this). These temperatures have increased as a result of anthropogenically driven climate change [@gruber]. Warming is shown to slowly move down into the water column thus increasing thermal stratification (RWS: citation needed). This thermal stratification restricts the depth at which water is upwelled and therfore limits the amount of nutrients brought up to the surface with a decrease in oxygen and pH levels [@chhak; @jacox2011; @jocox2016]. Climate change impacting oceanic processes has major ecological consequences [@walther]. This is demonstrated in several regions increases in sea surface temperatures have been proven to greatly influence marine species survival, reproduction and abundance. Climate change has also been shown to largely influence coastal species such as corals, muscles and oysters [@hoegh]. Given the importance of primary production, it was determined that the Benguela upwelling system, the eastern boundary upwelling system in South Africa, should be the focus of this paper. Furthermore, it should be understood that increases in sea surface temperatures bear considerable economic impacts as well [@murawski2000; @murawski1993] therefore understanding the impact and cause of climate change has merit not only in mitigation efforts in limiting environmental degradation and maintaining ecosystem functioning, but in its negative affects to local and broader economies as well as human populations dependent on these coastal regions.

The 3000km South African coastline is bordered by the Agulhas current along the east coast and the Benguela current (an EBUS) along the west coast. These coastal regions display noticeable differences in seawater temperatures primarily due to the influences of the neighbouring ocean currents [@Mead; @schlegel; @schlegel2016]. The Benguela Current is driven by the anticyclone high pressure systems and eastward moving cyclones which determine the boundaries of the Benguela upwelling region [@jarred; @hutchings]. The atmospheric temperatures along the coastline are largely influenced by the Benguela Current [@van], however, upwelling systems may be forced by various atmospheric conditions [@Lutjeharms; @roberts; @hutchings]. Regionally, the Benguela Current assists in transporting cold water northwards from the Southern Ocean to the coast [@luning1990; @lutjeharms2000; @hutchings; @schlegel].

The Benguela system is bounded by the equatorward and poleward ends by warm water regimes. These currents are known to be forced by local wind stress fields off southwest Africa [@nelson; @fennel]. This particular EBUS stretches along the south-western coast of Africa, from southern Angola to Cape Agulhas [@cole], with water temperatures coldest and most persistent off the Namibian coast at 25°S and weakest at 17°S [@Lutjeharms]. The upwelling period in the Benguela north of 25°C persists throughout the entire year but peaks during August. The Benguela upwelling region is split into northern and southern systems by a zone of intense perennial upwelling activity in the Luderitz region in Namibia [@shannon; @cole]. Several studies have shown that extreme warming and cooling periods occur in the northern and southern Benguela systems [@stander; @walker; @shannon1984; @boyd; @brainard]. Furthermore, Walter (1937) found that warm periods of global ocean temperatures may have resulted as a lack of upwelling or poor upwelling events. During the periods of warming, the equatorward wind stress over the northern region of the Benguela system was stronger than usual yet upwelling was suppressed by the southward movement of warm saline waters from the Angolan region. In the southern Benguela region, continental shelf bathymetry and strong upwelling favourable winds are known to result in more intense upwelling events. Upwelling occurring along the southern coastline of southern Africa are usually generated by local winds.

@@

Strong temperature gradients existing between offshore and inshore oceanic waters may result in different upwelling event readings. Upwelling relatively close to the shore has been found to depend on the strength and direction of the wind and tend to vary based on seasonal wind variation [@andrews]. Inshore temperatures tend to be lower during summer months due to upwelling, but there is an increase in offshore temperatures in the later summer months as a result of warming of the South East Atlantic Surface waters [@andrews; @Shannon]. Due to the temperature gradients that exist inshore and offshore, this study focused on observing upwelling patterns from the coastal region to 50km from the coastline.

The central assumption of this project was that remotely-sensed SST data and *in situ* seawater temperature data may be compared. Remotely sensed SST data and *in situ* seawater temperature data were obtained at different resolutions and so slight changes may be found within the different datasets. The primary aim of the first study was to make use of various remotely sensed SST and *in-situ* collected coastal seawater temperature data. The most appropriate dataset for the measurement of the nearshore temperatures within South Africa was the South African Coastal Temperature Network (SACTN) dataset [@smit]. It was useful to examine whether or not the same upwelling patterns exist at different distances from the coastline and if these patterns that exist are evident within the various datasets. This investigation examined recent trends and variation of upwelling events at different distances from the coastline.

@@

The South African coastline exhibits a large variation in seawater temperature and is divided into four bioregions [@smit, @mead]. The western region of the coastline is dominated by the Benguela Current forming an Eastern Boundry Upwelling System (EBUS) [@hutchings], which provides a natural laboratory for this study. Annual mean coastal seawater temperatures within this region have a range of 12.3 ± 1.2°C at the Western limit near the Namibian border. Seasonal upwelling is controlled by south-easterly trade winds and as a result reflects distinct temperature variations representing much lower temperatures within the upwelling cells over a fairly narrow continental shelf found from the Cape Peninsula to Cape Columbine. In order to accurately examine upwelling patterns at various distances from the coast, several sites from the SACTN dataset along the west coast of South Africa (i.e. sites influenced by the Benguela current) were selected. Since temperature data were not evenly recorded for each of the sites, the entire dataset was narrowed to only those sites with a time series of longer than 30 years. This yielded a total of four sites within the Benguela region (Fig 1).

@@

This study aimed to investigate patterns occurring within upwelling events at various different spatial intervals from the coastline, as well as to test whether the same patterns are discernible in the various datasets. As seawater temperature is known to have a large influence on species distribution and productivity [@Bolton2010; @smit2013], thus it is important to understand not only how temperatures vary at different distances from the coastline but what contributing factors to ecosystem maintenance the intensity of these events have across these distance intervals from the coastline. Of particular interest to this study; the Benguela upwelling system, which is distinct in that it is one of four most productive regions in the world.

@@

We would be amiss to point out some weaknesses inherent in the SACTN dataset. Temperature measurement along the South African coastline started in the 1970s, and has been inconsistent in terms of the instruments used, the modifications to styles and methods, calibration, as well as site locations themselves [@Smit2013; @schlegel2016]. Additionally, older records within some datasets, such as the SACTN dataset, may have been lost or are unreliable since the metadata for these are no longer available. These metadata are essential as their absence prevents the understanding of the influence regarding the utilised instruments on temperature recordings. It can thus be speculated that some measurements may not accurately represent well-validated temperature data due to the instrumentation used; however, we mitigated these effects by selecting only the most reliable and longest duration data, which other current and on-going studies, such as that by Williamson et al. have suggested are most suitable for the problem at hand.

>>>>

Upwelling is a physical oceanographic process in which cold, nutrient rich waters are pushed to the surface as a result of the motion caused by alongshore equatorial winds. These upwelling events are known to have a large influence on the abundance, diversity, and distribution of marine organisms, and moreover, critically affect marine ecosystem processes (Add in references). Occurring globally throughout the world's oceans, four major eastern boundary upwelling systems (EBUS) can be recognised; each significantly impacting associated coastal systems. These respective systems are characterized as vast regions of coastal ocean occurring along the western shores of continents bordering the Pacific and Atlantic Oceans [@bakun; @pauly]. There has been significant interest in major upwelling systems since the later 1960s, with major interest occuring off Peru [@walsh] and North-West Africa [@barber]. Research done by Copenhagen 1953, reported a number of sites along the west coast of southern Africa where upwelling was particularly intense. The two sites of intense upwelling activity were found off the Cape Peninsula and off Cape columbine [@copenhagen; @andrews]. Whilst upwelling plays a significant role in ecological processes, climate change largely influences this phenomenon.

@@

Several studies have provided significant evidence suggesting that there has been an increase in ocean surface temperatures over the last few decades (Add references; RWS: You can cite the IPCC for this). These temperatures have increased as a result of anthropogenically driven climate change [@gruber]. Warming is shown to slowly move down into the water column thus increasing thermal stratification (RWS: citation needed). This thermal stratification restricts the depth at which water is upwelled and therfore limits the amount of nutrients brought up to the surface with a decrease in oxygen and pH levels [@chhak; @jacox2011; @jacox2016]. Climate change impacting oceanic processes has major ecological consequences [@walther]. This is demonstrated in several regions increases in sea surface temperatures have been proven to greatly influence marine species survival, reproduction and abundance. Climate change has also been shown to largely influence coastal species such as corals, muscles and oysters [@hoegh]. Given the importance of primary production, it was determined that the Benguela upwelling system, the eastern boundary upwelling system in South Africa, should be the focus of this paper. Furthermore, it should be understood that increases in sea surface temperatures bear considerable economic impacts as well [@murawski2000; @murawski1993] therefore understanding the impact and cause of climate change has merit not only in mitigation efforts in limiting environmental degradation and maintaining ecosystem functioning, but in its negative affects to local and broader economies as well as human populations dependent on these coastal regions.

The 3000km South African coastline is bordered by the Agulhas current along the east coast and the Benguela current (an EBUS) along the west coast. These coastal regions display noticeable differences in seawater temperatures primarily due to the influences of the neighbouring ocean currents [@mead; @schlegel; @schlegel2016]. The Benguela Current is driven by the anticyclone high pressure systems and eastward moving cyclones which determine the boundaries of the Benguela upwelling region [@jarre; @hutchings]. The atmospheric temperatures along the coastline are largely influenced by the Benguela Current [@van], however, upwelling systems may be forced by various atmospheric conditions [@lutjeharms1987; @lutjeharms2003; @roberts; @hutchings]. Regionally, the Benguela Current assists in transporting cold water northwards from the Southern Ocean to the coast [@luning1990; @lutjeharms2000; @hutchings; @schlegel].

The Benguela system is bounded by the equatorward and poleward ends by warm water regimes. These currents are known to be forced by local wind stress fields off southwest Africa [@nelson; @fennel]. This particular EBUS stretches along the south-western coast of Africa, from southern Angola to Cape Agulhas [@cole], with water temperatures coldest and most persistent off the Namibian coast at 25°S and weakest at 17°S [@lutjeharms2000]. The upwelling period in the Benguela north of 25°C persists throughout the entire year but peaks during August. The Benguela upwelling region is split into northern and southern systems by a zone of intense perennial upwelling activity in the Luderitz region in Namibia [@shannon; @cole]. Several studies have shown that extreme warming and cooling periods occur in the northern and southern Benguela systems [@stander; @walker; @shannon1984; @boyd; @brainard]. Furthermore, Walter (1937) found that warm periods of global ocean temperatures may have resulted as a lack of upwelling or poor upwelling events. During the periods of warming, the equatorward wind stress over the northern region of the Benguela system was stronger than usual yet upwelling was suppressed by the southward movement of warm saline waters from the Angolan region. In the southern Benguela region, continental shelf bathymetry and strong upwelling favourable winds are known to result in more intense upwelling events. Upwelling occurring along the southern coastline of southern Africa are usually generated by local winds.

@@

Strong temperature gradients existing between offshore and inshore oceanic waters may result in different upwelling event readings. Upwelling relatively close to the shore has been found to depend on the strength and direction of the wind and tend to vary based on seasonal wind variation [@andrews]. Inshore temperatures tend to be lower during summer months due to upwelling, but there is an increase in offshore temperatures in the later summer months as a result of warming of the South East Atlantic Surface waters [@andrews; @shannon]. Due to the temperature gradients that exist inshore and offshore, this study focused on observing upwelling patterns from the coastal region to 50km from the coastline.

The central assumption of this project was that remotely-sensed SST data and *in situ* seawater temperature data may be compared. Remotely sensed SST data and *in situ* seawater temperature data were obtained at different resolutions and so slight changes may be found within the different datasets. The primary aim of the first study was to make use of various remotely sensed SST and *in-situ* collected coastal seawater temperature data. The most appropriate dataset for the measurement of the nearshore temperatures within South Africa was the South African Coastal Temperature Network (SACTN) dataset [@smit2013]. It was useful to examine whether or not the same upwelling patterns exist at different distances from the coastline and if these patterns that exist are evident within the various datasets. This investigation examined recent trends and variation of upwelling events at different distances from the coastline.

@@

The South African coastline exhibits a large variation in seawater temperature and is divided into four bioregions [@smit2013, @mead]. The western region of the coastline is dominated by the Benguela Current forming an Eastern Boundry Upwelling System (EBUS) [@hutchings], which provides a natural laboratory for this study. Annual mean coastal seawater temperatures within this region have a range of 12.3 ± 1.2°C at the Western limit near the Namibian border. Seasonal upwelling is controlled by south-easterly trade winds and as a result reflects distinct temperature variations representing much lower temperatures within the upwelling cells over a fairly narrow continental shelf found from the Cape Peninsula to Cape Columbine. In order to accurately examine upwelling patterns at various distances from the coast, several sites from the SACTN dataset along the west coast of South Africa (i.e. sites influenced by the Benguela current) were selected. Since temperature data were not evenly recorded for each of the sites, the entire dataset was narrowed to only those sites with a time series of longer than 30 years. This yielded a total of four sites within the Benguela region (Fig 1).

@@

This study aimed to investigate patterns occurring within upwelling events at various different spatial intervals from the coastline, as well as to test whether the same patterns are discernible in the various datasets. As seawater temperature is known to have a large influence on species distribution and productivity [@bolton2010; @smit2013], thus it is important to understand not only how temperatures vary at different distances from the coastline but what contributing factors to ecosystem maintenance the intensity of these events have across these distance intervals from the coastline. Of particular interest to this study; the Benguela upwelling system, which is distinct in that it is one of four most productive regions in the world.

@@

We would be amiss to point out some weaknesses inherent in the SACTN dataset. Temperature measurement along the South African coastline started in the 1970s, and has been inconsistent in terms of the instruments used, the modifications to styles and methods, calibration, as well as site locations themselves [@smit2013; @schlegel2016]. Additionally, older records within some datasets, such as the SACTN dataset, may have been lost or are unreliable since the metadata for these are no longer available. These metadata are essential as their absence prevents the understanding of the influence regarding the utilised instruments on temperature recordings. It can thus be speculated that some measurements may not accurately represent well-validated temperature data due to the instrumentation used; however, we mitigated these effects by selecting only the most reliable and longest duration data, which other current and on-going studies, such as that by Williamson et al. have suggested are most suitable for the problem at hand.
